---
title: "Brazil / Europe ANS Performance Dashboard"
subtitle: "User Guide and Data Roadmap"
author: "DECEA / EUROCONTROL PBWG"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    toc-title: "Contents"
    number-sections: true
    theme: cosmo
    code-fold: true
    embed-resources: true
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(dplyr)
library(tidyr)
library(ggplot2)
library(knitr)
library(kableExtra)
```

# Introduction

This document describes the **Brazil / Europe ANS Performance Dashboard**, an
interactive R Shiny application for comparing Air Navigation System (ANS)
operational performance between Brazil (DECEA) and Europe (EUROCONTROL).

The dashboard is based on the joint report
*"Comparison of Operational Air Navigation System Performance: Brazil / Europe,
2019–2025"*, produced by the **Performance Benchmarking Working Group (PBWG)**,
and is designed to support the **data acquisition and monitoring process**
between the two regions by making the report's indicators accessible in an
interactive and updatable format.

## Background

In 2025, Brazil handled approximately **2.1 million controlled flights**,
representing around 19% of European traffic over the same period. The scale
difference — with Europe operating at a traffic density roughly four times
higher than Brazil — is one of the key structural characteristics explaining
the performance variations documented throughout the report.

The dashboard allows users to visualise these differences dynamically, compare
individual study airports, and track the evolution of each indicator over time.

# Running the Dashboard

## Prerequisites

- **R** version ≥ 4.1
- **RStudio** (recommended) or any R terminal

## Installation

```r
# Run once to install all required packages
source("setup.R")
```

The packages installed are:

| Package | Role in the dashboard |
|---------|-----------------------|
| `shiny` | Web application framework |
| `bslib` | Modern theme and layout (Bootstrap 5) |
| `dplyr` / `tidyr` | Data manipulation and reshaping |
| `ggplot2` | All charts and plots |
| `lubridate` | Date parsing |
| `scales` | Axis formatting |
| `DT` | Interactive data tables |
| `zoo` | Rolling averages for time series |

## Launching the Dashboard

```r
shiny::runApp()
```

The dashboard opens automatically in the default browser.

# Dashboard Structure

The dashboard is organised into **8 tabs**, each focused on a specific
performance area.

## Overview {#sec-overview}

**Purpose:** executive summary with the main figures and general traffic trend.

Contents:

- **KPI cards** — total controlled flights in 2025 for Brazil and Europe,
  number of study airports
- **Daily traffic chart** — 7-day rolling average of daily flights for both
  regions (2023–2025)
- **DAIO distribution** — percentage share of each operation type
  (Domestic/Regional, Departures, Arrivals, Overflights) in 2025

## Traffic {#sec-traffic}

**Purpose:** analysis of traffic volume with daily granularity and segmentation
by operation type.

Available filters:

- Region (BRA, EUR, or both)
- Flight segment (Total, Domestic/Regional, Departures, Arrivals, Overflights)

Contents:

- Daily time-series chart with 7-day rolling average
- Annual totals table by region

## Punctuality {#sec-punctuality}

**Purpose:** analysis of arrival and departure punctuality at the 24 study
airports, broken down into five time bands.

The **five punctuality bands** follow the PBWG report definition:

| Band | Definition |
|------|------------|
| Early > 15 min | Arrived/departed more than 15 min ahead of schedule |
| Early 5–15 min | Arrived/departed 5 to 15 min ahead of schedule |
| Within ± 5 min | Within the on-time window |
| Late 5–15 min | Arrived/departed 5 to 15 min behind schedule |
| Late > 15 min | Arrived/departed more than 15 min behind schedule |

Available filters:

- Phase (Arrival / Departure)
- Year
- Region (Brazil, Europe, or both)

Contents:

- Stacked bar chart with percentage distribution per airport
- % within ±15 min bar charts, one panel per region

> **Report highlight:** in 2025, departure punctuality in Brazil reached 81%
> of flights within the ±15 min window, compared with 66% in Europe.
> Brazilian airports also show a high share of "significantly early" arrivals
> (29%), suggesting conservative schedule buffering by operators.

## Capacity {#sec-capacity}

**Purpose:** analysis of declared capacity and utilisation indices for the
study airports over 2019–2025.

Indicators shown:

| Indicator | Definition |
|-----------|------------|
| `MAX_CAP` | Declared peak capacity (movements/hour) |
| `BLI` | *Busy-Level Index* — share of hours with demand > 20% of capacity |
| `PLI` | *Peak-Level Index* — share of hours with demand > 80% of capacity |

Available filters:

- Region
- Year range

Contents:

- Capacity ranking by airport (latest year)
- BLI and PLI evolution over time

## Taxi & ASMA {#sec-taxi}

**Purpose:** monitoring of additional surface times — key tactical efficiency
indicators for ground operations and arrival sequencing.

Indicators are calculated as the **deviation from the unimpeded reference time**
(the time a flight would take with no restrictions):

| Indicator | Phase | Definition |
|-----------|-------|------------|
| Additional Taxi-Out | Departure (DEP) | Extra time from gate pushback to take-off |
| Additional Taxi-In | Arrival (ARR) | Extra time from landing to gate arrival |
| Additional ASMA | Arrival | Extra time in the sequencing area (40 NM from airport) |

> **Note:** ASMA data is currently available for European airports only.

Available filters:

- Phase (Taxi-Out / Taxi-In)
- Region

## BRA-EUR Flows {#sec-flows}

**Purpose:** analysis of inter-regional connections between Brazil and Europe.

Contents:

- **Route pair ranking** — rank evolution of study airport pairs between
  the two regions (2023–2025), identifying which routes gained or lost
  importance
- **World region distribution** — share of each world region in the
  international departure traffic of Brazil and Europe in 2025

> **Report highlight:** in 2025, Madrid–Guarulhos (LEMD–SBGR) became the
> busiest route pair between the two regions, overtaking Lisbon–Guarulhos
> (LPPT–SBGR), which had led in 2024.

## Airport Comparison {#sec-comparison}

**Purpose:** direct, user-selected comparison between one Brazilian airport
and one European airport.

Contents:

- Punctuality distribution (arrival or departure) for each selected airport
  in the chosen year
- **Ordered throughput curve** — hourly arrival counts ranked from highest
  to lowest, showing the distribution of operational pressure across years
  (available for SBGR and LPPT, 2019–2025)

## Data {#sec-data}

**Purpose:** direct access to all loaded data tables for inspection, filtering,
and export.

Available tables: Airports, Traffic daily, Punctuality, BLI/PLI,
Taxi/ASMA, Flow pairs.

# Study Airports

```{r airports-table}
airports <- tribble(
  ~ICAO, ~Name, ~City, ~Region,
  "SBBR","Brasília International","Brasília","Brazil",
  "SBCF","Confins","Belo Horizonte","Brazil",
  "SBCT","Afonso Pena","Curitiba","Brazil",
  "SBEG","Eduardo Gomes","Manaus","Brazil",
  "SBGL","Galeão","Rio de Janeiro","Brazil",
  "SBGR","Guarulhos","São Paulo","Brazil",
  "SBKP","Viracopos","Campinas","Brazil",
  "SBPA","Salgado Filho","Porto Alegre","Brazil",
  "SBRF","Guararapes","Recife","Brazil",
  "SBRJ","Santos Dumont","Rio de Janeiro","Brazil",
  "SBSP","Congonhas","São Paulo","Brazil",
  "SBSV","Dep. Luís Eduardo Magalhães","Salvador","Brazil",
  "EDDF","Frankfurt Airport","Frankfurt","Europe",
  "EDDM","Munich Airport","Munich","Europe",
  "EGKK","Gatwick Airport","London","Europe",
  "EGLL","Heathrow Airport","London","Europe",
  "EHAM","Amsterdam Schiphol","Amsterdam","Europe",
  "LEBL","El Prat","Barcelona","Europe",
  "LEMD","Barajas","Madrid","Europe",
  "LFPG","Charles de Gaulle","Paris","Europe",
  "LGAV","Eleftherios Venizelos","Athens","Europe",
  "LPPT","Humberto Delgado","Lisbon","Europe",
  "LSZH","Kloten","Zurich","Europe",
  "LTFM","Istanbul Airport","Istanbul","Europe"
)

airports |>
  kbl(caption = "24 study airports from the PBWG joint report") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                full_width = FALSE) |>
  pack_rows("Brazil", 1, 12) |>
  pack_rows("Europe", 13, 24)
```

# Data Sources

All files in `data/` were produced by the **PBWG (Performance Benchmarking
Working Group)** from official DECEA and EUROCONTROL/PRU databases, and used
in the 2019–2025 edition of the joint report.

```{r data-sources-table}
tibble::tribble(
  ~File, ~Content, ~Period, ~Tab,
  "PBWG-BRA-network-traffic-2023-2025.csv",
    "Daily flight volume BRA by segment", "2023–2025", "Overview / Traffic",
  "PBWG-EUR-network-traffic-2023-2025.csv",
    "Daily flight volume EUR by segment", "2023–2025", "Overview / Traffic",
  "PBWG-BRA-EUR-daio-share-2025.csv",
    "DAIO distribution by region 2025", "2025", "Overview",
  "PBWG-BRA-punc-2025.csv",
    "Daily BRA punctuality by airport and phase", "2025", "Punctuality",
  "PBWG-EUR-punc-2025.csv",
    "Daily EUR punctuality by airport and phase", "2025", "Punctuality",
  "PBWG-EUR-PUNC-2024.csv",
    "EUR punctuality 2024", "2024", "Punctuality",
  "PBWG-EUR-PUNC-2023.csv",
    "EUR punctuality 2023", "2023", "Punctuality",
  "PBWG-EUR-LGAV-punc-2024.csv",
    "LGAV (Athens) punctuality 2024 — supplementary", "2024", "Punctuality",
  "PBWG-EUR-PUNC-LPPT-2019-2024.csv",
    "Historical LPPT (Lisbon) punctuality 2019–2024", "2019–2024", "Punctuality / Comparison",
  "PBWG-BRA-EUR-bli-pli-2019-2025.csv",
    "Declared capacity, BLI and PLI per airport", "2019–2025", "Capacity",
  "PBWG-BRA-txxt-analytic-[year]-ref2024-icao_ganp_p20.csv",
    "Additional taxi times BRA (3 files: 2023, 2024, 2025)", "2023–2025", "Taxi & ASMA",
  "PBWG-EUR-txxt-analytic-[year]-ref2024-icao_ganp_p20.csv",
    "Additional taxi times EUR (3 files: 2023, 2024, 2025)", "2023–2025", "Taxi & ASMA",
  "PBWG-EUR-asma40-monthly-2023-2025-public.csv",
    "Monthly additional ASMA time EUR (40 NM radius)", "2023–2025", "Taxi & ASMA",
  "PBWG-BRA-EUR-study-flow-rank-2023-2025.csv",
    "Route pair rankings between study airports", "2023–2025", "BRA-EUR Flows",
  "PBWG-BRA-EUR-world-region-departures-2025.csv",
    "International departure distribution by world region", "2025", "BRA-EUR Flows",
  "PBWG-BRA-EUR-ordered-throughput-LPPT-SBGR-2019-2025.csv",
    "Ordered hourly throughput curve (SBGR and LPPT)", "2019–2025", "Airport Comparison"
) |>
  kbl(caption = "Data files and corresponding dashboard tabs") |>
  kable_styling(bootstrap_options = c("striped","hover","condensed"),
                font_size = 12) |>
  column_spec(1, monospace = TRUE, width = "30em")
```

## Punctuality File Format

The punctuality files use the **ICAO GANP minute-bucket format**. Each row
represents one airport × date × phase combination, with flight counts per
minute-deviation band relative to the scheduled time:

```
ICAO, DATE, PHASE, N_VALID,
(-INF,-60], (-60,-55], ..., (-5,0], (0,5), ..., [55,60), [60,INF)
```

Negative values indicate early operations; positive values indicate late
operations. The dashboard aggregates these 28 buckets into the five standard
PBWG bands at load time.

# Publishing the Dashboard Online

The dashboard is designed to be published on
**[shinyapps.io](https://www.shinyapps.io)**, RStudio's hosting platform
for Shiny applications.

## Platform

| Feature | Free tier |
|---------|-----------|
| Number of apps | 5 |
| Active hours/month | 25 h |
| App URL | `https://<account>.shinyapps.io/<app-name>` |
| R version support | Matches your local R version |

The free tier is sufficient for internal use, stakeholder reviews, and
bilateral presentations between DECEA and EUROCONTROL.

## How to Deploy

### Step 1 — Create an account

Go to <https://www.shinyapps.io> and create a free account.

### Step 2 — Get your credentials

In the shinyapps.io dashboard:
`Account → Tokens → Add Token → Show → Copy to Clipboard`

### Step 3 — Fill in `deploy.R`

Open `deploy.R` in RStudio and replace the placeholder values:

```r
rsconnect::setAccountInfo(
  name   = "your-account-name",
  token  = "your-token",
  secret = "your-secret"
)
```

### Step 4 — Run

```r
source("deploy.R")
```

The console prints the public URL and opens the browser automatically.

## Security Note

`deploy.R` is safe to commit to GitHub — credentials are stored locally by
`rsconnect` after `setAccountInfo()` is called. The `rsconnect/` folder
(which holds the actual token files) is listed in `.gitignore` and will
never be pushed to the repository.

## Re-deploying After Updates

Every time data or code changes, re-running `source("deploy.R")` pushes
the new version to the same URL. No manual steps required.

---

# Data Update Roadmap

## Overview

The CSV files currently in `data/` were generated manually from DECEA and
EUROCONTROL databases to produce the 2019–2025 report edition.

The next step is to **automate** the generation of these files so the dashboard
can be refreshed without manual intervention each time new data becomes available.

## Planned Architecture

```
Original data sources
(DECEA / EUROCONTROL PRU)
         │
         ▼
  R extraction and
  processing scripts
  (to be added)
         │
         ▼
  data/*.csv  ←── Files read by the dashboard
         │
         ▼
  app.R (Shiny Dashboard)
```

## Planned Scripts

The table below lists the R scripts that will be added to the repository to
regenerate each dataset:

| Script (planned) | Data generated | Status |
|------------------|---------------|--------|
| `R/build_traffic.R` | Daily flight volume BRA and EUR | 🔲 Planned |
| `R/build_punctuality.R` | Punctuality distributions (minute buckets) | 🔲 Planned |
| `R/build_taxi_asma.R` | Additional taxi and ASMA times | 🔲 Planned |
| `R/build_capacity.R` | Declared capacity, BLI and PLI | 🔲 Planned |
| `R/build_flows.R` | BRA-EUR route pairs and world region distribution | 🔲 Planned |
| `R/build_throughput.R` | Ordered throughput curves by airport | 🔲 Planned |
| `R/run_all.R` | Master script that runs all of the above in sequence | 🔲 Planned |

## File Naming Convention

Data files follow the pattern:

```
PBWG-{REGION}-{indicator}-{period}-{reference}.csv
```

Examples:

- `PBWG-BRA-punc-2025.csv`
- `PBWG-EUR-txxt-analytic-2025-ref2024-icao_ganp_p20.csv`
- `PBWG-BRA-EUR-bli-pli-2019-2025.csv`

Update scripts must maintain this convention so the dashboard recognises
the files automatically.

## Next Steps

1. **Define data access** — establish connections to DECEA and EUROCONTROL/PRU
   databases
2. **Develop extraction scripts** — one script per indicator, following the
   `R/build_*.R` structure
3. **Test integration** — ensure that script-generated files match the exact
   schema of the current files
4. **Automate execution** — schedule periodic runs (monthly or after each data
   publication) via `R/run_all.R`
5. **Expand coverage** — add ASMA data for Brazil, historical BRA punctuality
   for years prior to 2025, and new indicators (HFE, VFE)

---

# Technical Notes

## Version Requirements

```r
R        >= 4.1
shiny    >= 1.8
bslib    >= 0.7
ggplot2  >= 3.5
dplyr    >= 1.1
```

## Performance

The app processes data files totalling approximately 800 MB of CSV on initial
load. On machines with ≥ 8 GB RAM the start-up time is around **15–30 seconds**.
Once loaded, all interactions are in-memory and respond in under 1 second.

To reduce load time in future versions, pre-processed data can be saved in
`.rds` format (R native), which is read approximately 10× faster than CSV.

## Reference

> DECEA / EUROCONTROL Performance Review Unit.
> *Comparison of Operational Air Navigation System Performance:
> Brazil / Europe, 2019–2025.*
> Performance Benchmarking Working Group (PBWG), 2025.
